---
title: 'Compendium Structural Equation Modeling in Educational Research'
author: "Suzanne Jak and Terrence D. Jorgensen"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preface {-}


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!--chapter:end:index.Rmd-->


# Installing `lavaan`

Placeholder


## Raw Data
## Summary Statistics from Raw Data
### Summary Statistics for Multiple Groups
## Importing Summary Data
### Full Covariance Matrix
#### Rescaling a Correlation Matrix
### Upper/Lower Triangle of a Covariance Matrix
#### Upper/Lower Triangle of a Correlation Matrix
## Linear Regression Models
## Matrix Notation
## Interpreting and Estimating Coefficients
## Syntax Examples with Raw Data
## Categorical Predictors
### Regression with a Baseline Group
### Regression with Group-Specific Intercepts
### Baseline Comparison vs. Group-Specific Intercepts
### $t$ vs. $F$ Tests
## Constant vs. Varying Slopes
## Reimport Data
## Raw Data vs. Summary Statistics
## What Is Covariance?
## Interpreting Covariance
## Covariance Matrix
## OLS with Summary Statistics: $\beta = (\mathbf{X}^{'} \mathbf{X})^{-1} \mathbf{X}^{'} \mathbf{y}$
## Intercepts and Means
## Intercepts and Means
## What Are Mean and Covariance *Structures*?

<!--chapter:end:01-ReviewRegression.Rmd-->

